\hoffset=0em
\centerline{{\it FM} formatting commands.}
\bigskip

\centerline{\it Nroff/man~commands with \tt fm~-mu}
\medskip

\item{\tt .sp~(n)}	Skip 1 (or n) lines.
\item{\tt \\fI}		Start {\it underlining} of text (termcap "ul").
\item{\tt \\fB}		Start {\bf bold} text (termcap "md").
\item{\tt \\fS}		Start {\tt reversed} text (termcap "mr").
\item{\tt \\fG}		Start {\sl blinking} text (termcap "mb").
\item{\tt \\fR}		End {\it underlining} or other special text
			(termcap "me"/"ue").
\item{\tt \\fP}		Same as {\tt \\fR} (bug).
\item{\tt .nf}		Stop filling lines to maximum width.
\item{\tt .fi}		Resume filling lines to maximum width.
\item{\tt .I~<line>}	Display <line> {\it underlined}.
\item{\tt .B~<line>}	Display <line> in {\bf bold}.
\item{\tt .S~<line>}	Display <line> {\tt reversed}.
\item{\tt .G~<line>}	Display <line> {\sl blinking}.
\item{\tt .SM~<line>}	Display <line> {\tt reversed}.
\item{\tt .R~<line>}	Display <line> plain.
\item{\tt .I}
\item{\tt .ft I}	Display next line {\it underlined}.
\item{\tt .B}
\item{\tt .ft B}	Display next line {\bf bold}.
\item{\tt .R}
\item{\tt .ft R}
\item{\tt .ft}		Display next line plain.
\item{\tt .CT~<char>}	Display <char> reversed or "<CTRL/<char>>"
			in case neither -u or -o was given.
\item{\tt \\*(lq}	Display "`".
\item{\tt \\*(rq}	Display "'".
\item{\tt \\(*<char>}	Display <char> reversed.
\item{\tt \\(<char><char>}
			Display <char><char> reversed.
\item{\tt \\s(-)<digit>}
			Ignored.
\item{\tt \\| \\\^ \\\& \\! \\\{ \\\} \\a \\c \\d \\e \\p \\r \\t \\u \\z}
			Ignored.
\item{\tt \\}		Display following character (except as above).
\item{\tt .br}		Terminate preceding paragraph.
\item{\tt .hy~<digit>}  Turn hyphenation off with 0 or on with 1-9.
			(See fm.1 for the differences among levels 1-9.)
\item{\tt .PP}		Terminate preceding paragraph.
\item{\tt .P}		Terminate preceding paragraph.
\item{\tt .LP}		Terminate preceding paragraph.
\item{\tt .TP~(n)}	Begin hanging paragraph with 5 (or n) extra
			spaces of indentation at left, with first
			following line in left margin.
\item{\tt .IP~<word>~(0.3i)~...}
			Begin hanging paragraph with 5 extra
			spaces of indentation at left, with <word> ...
			in left margin.
\item{\tt .HP~(n)}	Begin hanging paragraph with 5 extra
			spaces of indentation at left.
\item{\tt .RS}		Indent 5 extra spaces until {\tt .RE}.
\item{\tt .RE}		Resume normal indent after {\tt .RS}.
\item{\tt .DS}		Skip a line; preserve lines and spacing;
			indent 5 extra spaces until {\tt .DE}.
\item{\tt .DE}		Resume filling and normal indent after {\tt .DS}.
\item{\tt .EX}		Skip a line; preserve lines and spacing;
			indent 5 extra spaces until {\tt .EE}.
\item{\tt .EE}		Skip a line; resume filling and normal indent
			after {\tt .EX}.
\item{\tt .VS~<line>}	Ignored.
\item{\tt .VE~<line>}	Ignored.
\item{\tt .NT~(<line>)}	Skip a line, indent 5 extra spaces, shorten
			line length by 5, display {\bf NOTE} bold and
			centered (or <line>), continue until {\tt .NE}.
\item{\tt .NE}		Resume normal indent and line length after
			{\tt .NT}.
\item{\tt .IR~<words>}	Join words alternating {\it underlined} and plain.
\item{\tt .BR~<words>}	Join words alternating {\bf bold} and plain.
\item{\tt .RI~<words>}	Join words alternating plain and {\it underlined}.
\item{\tt .BI~<words>}	Join words alternating {\bf bold} and {\it underlined}.
\item{\tt .RB~<words>}	Join words alternating plain and {\bf bold}.
\item{\tt .IB~<words>}	Join words alternating {\it underlined} and {\bf bold}.
\item{\tt .SH~<line>}	Skip a line, and display <line> left adjusted
			and {\bf bold}.
\item{\tt .SS~<line>}	Skip a line, and display <line> {\bf bold}.
\item{\tt .TH~<name>~<n>~(<text>)}
			Right adjust {\bf <name>(<n>)} (followed by <text>).
\item{\tt .UC~(<digit>)}
			Ignored.
\item{\tt .\\"~<line>}	Ignore <line>.
\item{\tt .PN~<line>}	Underline <line>.
\item{\tt .ta~<line>}	Ignore <line>.
\item{\tt .PD~<line>}	Ignore <line>.
\item{\tt .NX<cap>~<line>}
			Ignore <line>.
\item{\tt .DT}		Ignored.
\item{\tt .ti~+n~<line>}
			n extra spaces of indentation at left of <line>.
\item{\tt .MS~<word>~<n>}
			Display {\it <word>(<n>)}.
\item{\tt .if~n~<line>}	Process line.
\item{\tt .if~t~<line>} Do not process line.
\item{\tt .de~<line>}	Ignore line and next 1-4 lines until after a line
			starts with {\tt ..}.  (I.e. ignore nroff macro
			definition.)

\bigskip
\centerline{\it \TeX ~commands with \tt fm~-xu}
\medskip

\item{\tt \\hsize=<n>em}
			Assign maximum line width (1 em is
			one character width).  Default is 72 or the
			value after {\tt -} on the command line.
\item{\tt \\hoffset=<n>em}
			Assign horizontal offset (default is 4 or
			the value after {\tt -p} on the command line).
\item{\tt \\leftskip=<n>em}
			Leave n extra blanks to the left of each
			following line.  The effect is global, not local to
			a group.
\item{\tt \\rightskip=<n>em}
			Leave n extra blanks to the right of each
			following line.  The effect is global, not local to
			a group.
\item{\tt \\break}	Start new line.
\item{\tt \\par}	Start new paragraph.
\item{\tt \\parindent=<n>em}
			Assign paragraph indentation (default is
			5).  A negative value for hanging indent
			is accepted, but the sum with hoffset should
			not be negative.
\item{\tt \\parskip=<n>pt}
			Assign vertical skip made before each
			paragraph (default is 0).  The value given
			is rounded to the nearest number of whole
			lines, reckoning 12 points as the height of
			one line.  The effect is not local to a group.
\item{\tt \\noindent}	Don't indent this paragraph.
\item{\tt \\bigskip}	Skip a line.
\item{\tt \\medskip}	Skip a line.
\item{\tt \\smallskip}	Skip a line.
\item{\tt \\\$ \\\# \\\{ \\\} \\\& \\\% \\^ \\_ \\\~ \\' \\" \\\\}
			Display character following \\.
\item{\tt\\backslash}	Display \backslash.
\item{\tt \\it}		Begin {\it underlining} (termcap "ul").
\item{\tt \\sl}		Begin {\sl blinking} (termcap "mb").
\item{\tt \\bf}		Begin {\bf bold} (termcap "md").
\item{\tt \\rm}		Begin plain.
\item{\tt \\tt}		Begin {\tt reverse} (termcap "mr").
\item{\tt \\TeX}	Display \TeX.
\item{\tt \\centerline~\{<line>\}}
			Display <line> centered.
\item{\tt\\beginsection~<line>}
			Skip a line, display <line> in bold.
\item{\tt \\rightline~\{<line>\}}
			Display <line> right adjusted.
\item{\tt \\raggedright}
			Stop justifying (for current \{...\} group).
\item{\tt \\obeylines}	Don't join or split input lines
			(for current \{...\} group).
\item{\tt \\obeyspaces} Preserve spacing in input text
			(for current \{...\} group).
\item{\tt \\item\{...\}}
			Start hanging paragraph with 10 extra
			spaces of left indentation and ... displayed in
			left margin and indented 5.
\item{\tt \\itemitem\{...\}}
			Same as {\tt \\item}.
\item{\tt \\<space-character>}
			Break between words.
\item{\tt \~}		Space that does not break between words and
			is displayed with current display attribute.
\item{\tt \$}
\item{\tt \#}		Ignored.
\item{\tt \\bye}
\item{\tt \\end}	Terminate processing input file(s).

\bigskip
\rightline{Greg Lee, May, 1992}
\bye
